---
title: "Tidyverse Programming Extras"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Extra Recipes for Programming with the Tidyverse

This worksheet contains some examples of how you can deal with more
complex scenarios, when the simple solutions covered in the tutorial
don't work!

This worksheet was adapted from:
<https://github.com/ttimbers/prog-tidyverse-functions>

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
options(dplyr.summarise.inform = FALSE)
```

## What about passing multiple data variables to a function?

> #### 1. SOLUTION A - Pass the dots `(…)`

```{r pass_the_dots}
#The dots are the last input parameter in the function call
grouped_mean_dots <- function(data, summary_column, ...) {
  data %>%
    group_by(...) %>%
    summarise("mean_{{summary_column}}" := mean({{ summary_column }}, na.rm = T))
}

grouped_mean_dots(gapminder, pop, continent, year)

#Can also pass single inputs to (...)
grouped_mean_dots(gapminder, pop, continent)

# Careful with string input vs data variables
# What is this doing?
grouped_mean_dots(gapminder, pop, "continent", "year")
#c.f.
mean(gapminder$pop)
```

-   Note that using `(…)` perpetuates NSE/Data-masking to your bespoke
    function.

-   NOTE also that you can only pass one set of inputs to the (...) - so
    you can't use it to specify group variables *and* summary variables!

> #### 2. SOLUTION B - use `across()`

```{r multi_gps_across}
gap_means <- function(data, grouping_variables, summary_columns){
  data %>% 
    group_by(across({{ grouping_variables }})) %>% 
    summarise(across({{ summary_columns }}, ~ mean(., na.rm = T))) 
}

gap_means(gapminder, c(country, year), c(lifeExp, pop, gdpPercap))

#Or combined across() with (...) - here the dots need to be encased in c()
my_mean <- function(data, ...) {
  data %>% 
    summarise(across(c(...), ~ mean(.x, na.rm = TRUE)))
}

gapminder %>% my_mean(lifeExp, pop)
gapminder %>% my_mean(starts_with("p"))

```

#### For the above name the columns with `.names = ""`

use `.names = ""` when there are multiple variables (`:=` only works
with single variables)

```{r multi_gps_naming}
# Use the .names = "" option 
gap_mean_name_multicols <- function(data, grouping_variable, summary_columns) {
  data %>% 
    group_by(across({{grouping_variable}})) %>% 
    summarise(across({{ summary_columns }}, 
                     ~ mean(., na.rm = TRUE), 
                     .names = "mean_{.col}"))
}

gap_mean_name_multicols(gapminder, c(continent, year), c(lifeExp, pop, gdpPercap))

```

> #### 3. SOLUTION C - Use `all_of()/any_of()`for string vector input:
>
> [To allow a vector of strings as function input]{.underline}: Use
> `all_of()/any_of()` (or when you don't want to propagate data
> masking).
>
> When we have a vector of string inputs we can't use the `.data[[ ]]`
> trick we used for single string input. The `.data` pronoun can only be
> sub-setted with single column names. It doesn't support single-bracket
> indexing (i.e. .`data[ ]`).
>
> Instead we can use `all_of()` or `any_of()`.
>
> -   `all_of()` acts on all of the columns listed (they must all
>     exist).
>
> -   `any_of()` acts on all of the columns listed, if they exist, and
>     ignores missing ones
>
> -   Can use in combination with `across()` when passing string input
>     to a non-<tidy-select> function (e.g. `group_by`)

```{r strings_helper_funs}
slice_stringinputs_multgps <- function(data, grouping_variable, select_columns, row_start, row_end) {
  data %>% 
    group_by(across(all_of(grouping_variable))) %>% 
    select(all_of(select_columns)) %>% 
    slice(row_start:row_end)
}

slice_stringinputs_multgps(gapminder, c("continent", "country"), c("pop", "lifeExp"), 1, 3)

```

```{r strings_any_of_across}
slice_stringinputs_multgps <- function(data, grouping_variable, columns, row_start, row_end) {
  data %>% 
    group_by(across(any_of(grouping_variable))) %>% 
    select(any_of(columns)) %>% 
    slice(row_start:row_end)
}

slice_stringinputs_multgps(gapminder, c("continent", "country", "missing"), c("deleted", "pop", "lifeExp"), 1, 3)


## Again you need 'across' when using non-<tidy-select> functions
slice_stringinputs <- function(data, select_columns, rows) {
  data %>%
    group_by(across(all_of(select_columns))) %>%
    slice(rows)
}
slice_stringinputs(gapminder, c("country", "year"), 1)
```

------------------------------------------------------------------------

## **Unexpected Data Masking**

Unexpected data masking occurs when an object in your workspace (which
you want to use) has the same name as a dataset column.

Because tidyverse/dplyr verbs by default act on data set variables your
code may end up using the dataset variable not the environment variable.

```{r unexp_masking, eval = FALSE}
# Example of Unexpected masking
add_something <- function(data, col) {
  n <- sample.int(200, size=1)
  data %>% mutate(z = {{ col }} * y + n)
}

data1 <- data.frame(x = 100, y = 10)
data2 <- data.frame(x = 100, y = 10, n = "argh")

data1 %>% add_something(x) %>% pull(z)
data2 %>% add_something(x) %>% pull(z)

```

Here the data variable `n` in `data2` has unexpectedly masked our
environment variable `n`!

> #### SOLUTION - Be explicit in your code
>
> Use the `.data$` and `.env$` pronouns to explicitly associate
> variables to data frames or the global environment:

```{r unexp_mask_soln}
#Specify .data and .env pronouns
add_something <- function(data, col) {
  n <- sample.int(200, size=1)
  data %>% mutate(z = {{ col }} * .data$y + .env$n)
}

data1 <- data.frame(x = 100, y = 10)
data2 <- data.frame(x = 100, y = 10, n = "argh")

data1 %>% add_something(x) %>% pull(z)
data2 %>% add_something(x) %>% pull(z)
```

If you are making a package then the explicit use of the `.data$` and
`.env$` pronouns to define variables prevents warnings being issued when
you run the `check()` command when compiling the package.

------------------------------------------------------------------------

## Using tidy evaluation in ggplot

We can apply this same logic to other ggplot functions.

-   either embrace `{{ }}` the data variable names

```{r ggplot_NSE_functions}
scatter_plot <- function(data_frame, x_axis, y_axis ) {
  ggplot(data_frame, aes(x = {{ x_axis }}, y = {{ y_axis }})) +
    geom_point(alpha = 0.5)
}

gapminder %>%
  gap_mean_name_multicols(year, lifeExp) %>%
  scatter_plot(year, mean_lifeExp)

gapminder %>%
  gap_mean_name_multicols(continent, gdpPercap) %>%
  scatter_plot(continent, mean_gdpPercap)
```

-   or subset the data using `.data[[ ]]` and input strings to the
    function call

```{r}
scatter_plot <- function(data_frame, x_axis, y_axis ) {
  ggplot(data_frame, aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
    geom_point(alpha = 0.5)
}

gapminder %>%
  gap_mean_name_multicols(continent, gdpPercap) %>%
  scatter_plot("continent", "mean_gdpPercap")
```

-   or pass the dots `(…)` - can pass the dots to `aes()` or to `vars()`

```{r ggplot_pass_the_dots}
scatter_facet_plot <- function(data_frame, x_axis, y_axis, ...) {
  ggplot(data_frame, aes(x = {{ x_axis }}, y = {{ y_axis }})) +
    geom_point(alpha = 0.5) +
    facet_wrap(vars(...))
}

gapminder %>% 
  grouped_mean_dots(lifeExp, continent, year) %>% 
  scatter_facet_plot(x_axis = year, y_axis = mean_lifeExp, continent)
```

For more ggplot examples see:
`docs/ggplot_bespoke_functions_examples.docx`

------------------------------------------------------------------------

## Do you even need Tidy Eval?

#### Not if you can assume fixed column names in data!

-   Need to add checks on inputs (column existence, scale of values) in
    the function

```{r fixed_col_names}
# Calculate lifeExp / pop (a contrived example)

life_v_pop <- function(data){
  if(all(c("lifeExp", "pop") %in% names(data))) {
    data %>% mutate(life_v_pop = lifeExp / pop)
  
  } else {
    stop("dataset must contain columns named `lifeExp` and `pop`")
  }
}
  
life_v_pop(gapminder)
life_v_pop(mtcars)
```

#### Not if you can use column-wise mapping functions (scoped dplyr verbs)

-   Scoped verbs for `mutate_` / `summarise_` / `filter_` / `group_by_`

    -   `_all` --\> map over all columns

    -   `_if` --\> map over columns selected by condition

    -   `_at` --\> map over a custom selection

    -   No data masking = take objects (not blueprints)

```{r nonNSE_function_alternatives}
#Scale all numeric variables by SD (function will work on any dataset)
my_scale <- function(data){
  data %>% mutate_if(is.numeric, ~. / sd(.))
}

my_scale(gapminder)
my_scale(mtcars)

# Don't necessarily need to wrap in a function...
iris %>% mutate_if(is.numeric, ~. / sd(.))

#Another example -> mean of all numeric cols by all grouping variables
my_gp_summarise <- function(data){
  data %>% 
    group_by_if(is.factor) %>% 
    summarise_if(is.numeric, mean)
}

# works on any dataset
my_gp_summarise(gapminder)
my_gp_summarise(mtcars)

#Can use string vectors directly
my_summarise <- function(data, variables){
  data %>% summarise_at(variables, mean)
}

my_summarise(gapminder, c("lifeExp", "pop"))

```

------------------------------------------------------------------------

## Do we ever need for loops?

We might naively try to summarise two columns in a for loop:

```{r for_loop}
columns <- names(gapminder)[4:6]
out <- rep(list(NULL), 3)

for(col in seq_along(columns)){
  out[[col]] <- gapminder %>%
    summarise(mean = mean(columns[[col]], na.rm = T))
}

unlist(out)
```

It doesn't work because `columns[[col]]`, a string representing the name
of a column in gapminder, is not recognised as a data variable within
the context of the `mean` function. Now we can try again using the
`.data` pronoun.

```{r for_loop_solved}
columns <- names(gapminder)[4:6]
out <- rep(list(NULL), 3)

for(col in columns){
  out[[col]] <- gapminder %>%
    summarise(mean = mean(.data[[col]]))
}

unlist(out)
```

It worked but it's ugly and we have an awkward output format. We can
improve this by using the tidy selection function `across` to drop the
for loop:

```{r for_loop_replaced}
columns <- names(gapminder)[4:6]
gapminder %>%
  summarise(across(all_of(columns), mean))

# or, if we have parameters, use a dot to represent the current variable
gapminder %>%
    summarise(across(all_of(columns), ~ mean(., na.rm = T)))
```

> This actually means that for loops are rarely needed!

------------------------------------------------------------------------

## 
