---
title: "Tidyverse Programming Extras"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
```

## What about with multiple data variables?

> #### SOLUTION A - Pass the dots `(…)`
>
> Can only pass the dots once in a function! So only use when is one
> pipeline input.

```{r pass_the_dots}
#The dots are the last input parameter in the function call
grouped_mean_dots <- function(data, summary_column, ...) {
  data %>%
    group_by(...) %>%
    summarise("mean_{{summary_column}}" := mean({{ summary_column }}, na.rm = T))
}

grouped_mean_dots(gapminder, pop, continent, year)

# Careful with string input vs data variables
# What is this doing?
grouped_mean_dots(gapminder, pop, "year")
#c.f.
mean(gapminder$pop)
```

-   Note that using `(…)` perpetuates NSE/Data-masking to your bespoke
    function.

-   NOTE - pass the dots `(…)` is also very useful for single data
    variable functions and for ggplot (can pass the dots to `aes()` or
    `vars()`

```{r ggplot_pas_the_dots}
scatter_facet_plot <- function(data_frame, x_axis, y_axis, ...) {
  ggplot(data_frame, aes(x = {{ x_axis }}, y = {{ y_axis }})) +
    geom_point(alpha = 0.5) +
    facet_wrap(vars(...))
}

gapminder %>% 
  grouped_mean_dots(lifeExp, continent, year) %>% 
  scatter_facet_plot(x_axis = year, y_axis = mean_lifeExp, continent)

```

> #### SOLUTION B - use across()

```{r multi_gps_across}
gap_means <- function(data, grouping_variables, summary_columns){
  data %>% 
    group_by(across({{ grouping_variables }})) %>% 
    summarise(across({{ summary_columns }}, ~ mean(., na.rm = T))) 
}

gap_means(gapminder, c(country, year), c(lifeExp, pop, gdpPercap))

#Or combined across() with (...) - here the dots need to be encased in c()
my_mean <- function(data, ...) {
  data %>% 
    summarise(across(c(...), ~ mean(.x, na.rm = TRUE)))
}

gapminder %>% my_mean(lifeExp, pop)
gapminder %>% my_mean(starts_with("p"))

```

#### For the above - name the columns with `.names = ""`

use `.names = ""` when there are multiple variables (`:=` only works
with single variables)

```{r multi_gps_naming}
# Use the .names = "" option 
gap_mean_name_multicols <- function(data, grouping_variable, summary_columns) {
  data %>% 
    group_by(across({{grouping_variable}})) %>% 
    summarise(across({{ summary_columns }}, 
                     ~ mean(., na.rm = TRUE), 
                     .names = "mean_{.col}"))
}

gap_mean_name_multicols(gapminder, c(continent, year), c(lifeExp, pop, gdpPercap))

```

> #### SOLUTION C - for STRING INPUT can use  string vetor helpers, e.g. `all_of()` / `any_of()` or scoped dplyr verbs e.g. `summarise_at()` 
>
> Can't use the `.data[[ ]]` trick we used for single string input. The
> `.data` pronoun can only be subsetted with single column names. It
> doesn't support single-bracket indexing (i.e. .`data[ ]`):

```{r strings_helper_funs}
slice_character <- function(data, grouping_variable, columns, row_start, row_end) {
  data %>% 
    group_by(across(all_of(grouping_variable))) %>% 
    select(all_of(columns)) %>% 
    slice(row_start:row_end)
}
slice_character(gapminder, c("continent", "country"), c("pop", "lifeExp"), 1, 3)

```

```{r scoped_verbs_strings}
gap_means_multigps <- function(data, grouping_variables, summary_columns){
  data %>%
    group_by_at(vars(grouping_variables)) %>%
    summarise_at(vars(summary_columns), mean)
}
gap_means_multigps(gapminder, c("continent", "country"), c("pop", "lifeExp"))
```

## Do you even need Tidy Eval?

#### Not if you can assume fixed column names in data!

-   Need to add checks on inputs (column existence, scale of values) in
    the function

```{r fixed_col_names}
# Calculate lifeExp / pop (a contrived example)

life_v_pop <- function(data){
  if(all(c("lifeExp", "pop") %in% names(data))) {
    data %>% mutate(life_v_pop = lifeExp / pop)
  
  } else {
    stop("dataset must contain columns named `lifeExp` and `pop`")
  }
}
  
life_v_pop(gapminder)
life_v_pop(mtcars)
```

#### Not if you can use column-wise mapping functions (scoped dplyr verbs)

-   Scoped verbs for `mutate_` / `summarise_` / `filter_` / `group_by_`

    -   `_all` --\> map over all columns; `_if` --\> map over columns
        selected by condition; `_at` --\> map over a custom selection

    -   No data masking = take objects (not blueprints)

```{r map_function_alternatives}
#Function scales all numeric variables by SD (will work on any dataset)
gapminder %>% mutate_if(is.numeric, ~. / sd(.))
mtcars %>%  mutate_if(is.numeric, ~. / sd(.))

#Can use string vectors directly
gapminder %>% summarise_at(c("pop", "lifeExp"), mean)
gapminder %>% 
  group_by_if(is.factor) %>% 
  summarise_if(is.numeric, mean)

#work on any dataset
mtcars %>% 
  group_by_if(is.factor) %>% 
  summarise_if(is.numeric, mean)
```

------------------------------------------------------------------------
