---
title: "Tidyverse programming tutorial"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
options(dplyr.summarise.inform = FALSE)
```

# Outline

-   Motivation
    -   Why we love the tidyverse
    -   Where we start to get into difficulties
-   Background to programming, tidy-style
    -   Different types of variables
    -   Non-standard evaluation
    -   Tidy evaluation
-   Implementing tidy functions
    -   The curly-curly operator `{{ }}`
-   Data masking: problems and solutions
    -   Unexpected masking
    -   Naming variables
    -   String inputs
-   Tidy selection: problems and solutions
    -   Functions with arbitrary selection of variables

# Motivation

## ♥ Tidyverse = super easy to work with! ♥

Let's suppose we want to calculate the mean life expectancy per
continent or per year for the
[Gapminder](https://www.gapminder.org/data/documentation/ "Description of data from gapminder.org"){.uri}
dataset:

```{r contrived_example}
# Check what the dataset looks like
head(gapminder)

# Use tidyverse to calculate grouped means
gapminder %>%
  group_by(year) %>%
  summarise(meanLE = mean(lifeExp, na.rm = T))
gapminder %>% 
  group_by(continent) %>% 
  summarise(meanLE = mean(lifeExp, na.rm = T))
```

Simple!

Once we specify the dataset we are working with we can directly type the
names of the columns and variables in the dataset, without having to
enclose them in quotation marks or having to define them relative to the
dataset itself.

## But there's a problem! Trouble in paradise?

In the example above we calculated the same metric for two different
grouping variables by repeating the code and changing one element. This
isn't good coding practice. When we find ourselves repeating code we
should look at using some kind of iteration method or writing a bespoke
function.

Also, we might want to get the mean of population size and GDP per
capita as well as life expectancy across continents and years. Or we
might want to calculate it by countries AND years. We shouldn't repeat
the above code that many times! We need to write re-usable code.

SO let's see what happens when we try to turn the above code into a
function:

```{r try_function}
# Place code within function

group_mean <- function(data, grouping_variable, summary_column){
  data %>% 
    group_by(grouping_variable) %>% 
    summarise(mean = mean(summary_column, na.rm = T))
}

# Does it work?
group_mean(gapminder, year, lifeExp)

# Oh no! R does not understand what we mean. 
# What about if we place the arguments in quotes?
group_mean(gapminder, "year", "lifeExp")
# Still no. 

```

What now?

------------------------------------------------------------------------

# Background to programming, tidy-style

To write functions that use tidyverse functions, we need to

-   Understand **data masking**
-   Implement **non-standard evaluation**

### Data masking

The key idea behind data masking is that it blurs the line between the
two different meanings of the word *variable*:

-   **environment variables** are programming variables that live in an
    environment. They are often created with `<-` but they may also be
    function arguments, function names, or even built-in names such as
    `TRUE`.

-   **data variables** are the variables that are part of a particular
    data frame (the identifiers of the columns).

Data Masking allows you to use data variables as if they were
environment variables, i.e. it lets you write `my_variable` instead of
`dataset$my_variable`. This is used in tidyverse functions such as

-   `arrange()`, `count()`, `filter()`, `group_by()`, `mutate()`, and
    `summarise()`

Data masking makes data manipulation easier because it requires less
typing (cf with base R):

```{r data_masking, results = FALSE}
#Compare Base R
gapminder[gapminder$continent == "Europe" & gapminder$pop < 1000000, ]

#With Tidyverse
gapminder %>% filter(continent == "Europe", pop < 1000000)
```

This is (one of the reasons) why the tidyverse is so popular and easy to
use!

So how does the tidyverse do this? The answer lies in non-standard
evaluation.

### NSE = Non-standard evaluation (VERY VERY briefly!)

What is NSE? Very roughly, it is to "*programmatically modify an
expression or its meaning after it is issued but before it is
executed*". (You can think of an expression as an R command you might
type at the prompt or in an R script).

*Not many programming languages give the programmer the power to
implement, use, and abuse Non-Standard Evaluation - but R does!*

And not just in tidyverse - ALSO in base R!

```{r NSE_baseR}
subset(gapminder, continent=="Asia") # this works because subset delays the 
                                     # evaluation of continent=="Asia" 
                                     # until it is run when it changes its 
                                     # meaning/environment so it relates to gapminder
gapminder[continent=="Asia", ] # Does this work?
                               # NO! continent is not defined in our environment

#Other base R NSE functions include:
# with()
# transform()
# lm()
# and the old school (for those of us who learnt R way before the tidyverse!):
# attach()
```

Much more to learn about NSE!! See list of references at end.

### Tidy Evaluation

Tidy evaluation is a version of NSE used in the tidyverse (implemented
in the rlang package), designed for working with data. It provides:

**Data Masking** to access your data variables directly in the context
of the data (perhaps modified or grouped earlier in the pipeline) that
you are working with

**Tidy Selection** to easily choose variables based on their position,
name, or type, e.g. `starts_with("x")` or `is.numeric`, and write code
that may use one or many separate data variables together. This is used
in tidyverse selection functions such as

-   `across()`, `relocate()`, `rename_with()`, and `select()`

> Tidyverse functions are built for interactivity! They make interactive
> data exploration fast and fluid (and fun!)

> But they add some new challenges when you attempt to use them
> indirectly such as in a for loop or a function.

------------------------------------------------------------------------

# Implementing tidy functions

If you are writing a function that will call any tidyverse functions,
you are likely to want or need to make your function work in a similar
way.

## Standard tidyverse-style functions with curly-curly

In our group mean example above, we want our parameters
`grouping_variable` and `summary_column` to be treated as data
variables. To do this we will implement the function with tidy
evaluation!

Here is our example from above which didn't work:

    group_mean <- function(data, grouping_variable, summary_column) {
      data %>%
        group_by(grouping_variable) %>%
        summarise(mean = mean(summary_column, na.rm = T))
    }
    group_mean(gapminder, year, lifeExp)

Inside the function, `grouping_variable` and `summary_column` are
environment variables. But when you call `group_by` and `summarise`,
they expect the arguments to be data variables and look for columns
named `grouping_variable` and `summary_column` instead of `year` and
`lifeExp`.

Data-masked code needs to be *delayed* and *transported* to the data
context. Behind the scenes, dplyr verbs achieve this by capturing the
blueprint of your code, and resuming its evaluation inside the data
mask. The example above fails because `group_by()` is capturing the
wrong piece of blueprint.

> #### SOLUTION - use the curly-curly operator `{{ }}`
>
> The curly-curly `{{ }}` operator allows you to *tunnel* a data
> variable through a function. It acts as a wrapper for lower level
> operators (`enquo()` and `!!`) that function to **quote-and-unquote**
> the expression pattern: `enquo()` - delays the interpretation of code
> and captures its blueprint and `!!` modifies the blueprints.

```{r tidy_eval_soln}
group_mean <- function(data, grouping_variable, summary_column){
  data %>%
    group_by({{ grouping_variable }}) %>%
    summarise(mean({{ summary_column }}, na.rm = T))
}

#It works!
group_mean(gapminder, year, lifeExp)
group_mean(gapminder, continent, pop)

```

**Congratulations!** You have just written a function that uses tidy
evaluation! It was as simple as adding a few curly braces.

Of course, when you try to use this in real life you will encounter
plenty of difficulties along the way!

Here are some common problems and a few hints and tips.

## Data Masking Problems and Solutions

### **1. Unexpected masking**

```{r unexp_masking}
# Example of Unexpected masking
# An object in your workspace (which you want to use) has the 
# same name as a dataset column. 
n <- sample.int(200, size = 1)
add_something <- function(data, col) {
  data %>% mutate(z = {{ col }} * y + n)
}

data1 <- data.frame(x = 100, y = 10)
data2 <- data.frame(x = 100, y = 10, n = "argh")

data1 %>% add_something(x) %>% pull(z)
data2 %>% add_something(x) %>% pull(z)

```

Here the data variable `n` in `data2` has unexpectedly masked our
environment variable `n`!

> #### SOLUTION - Be explicit in your code
>
> Use the `.data$` or `.env$` pronouns to explicitly associate variables
> to data frames or the global environment:

```{r unexp_mask_soln}
#Specify .data and .env pronouns
n <- sample.int(200, size=1)
y <- NULL

add_something <- function(data, col) {
  data %>% mutate(z = {{ col }} * .data$y + .env$n)
}

data1 %>% add_something(x) %>% pull(y)
data2 %>% add_something(x) %>% pull(y)
```

### **2. Naming of variables in functions**

You may want to construct the name of a variable from some data or
environmental variables.

```{r name_assignment}
group_mean_name <- function(data, grouping_variable, summary_column) {
  result_name <- paste0("mean_", names(data %>% select( {{ summary_column }} )))
  #print the result_name to screen
  print(result_name)
  data %>%
    group_by({{ grouping_variable }}) %>%
    summarise(result_name = mean({{ summary_column }}, na.rm = T))
}

group_mean_name(gapminder, country, lifeExp)
# Doesn't work! The column has been named 'result_name'.
```

> #### Solution - use the walrus operator `:=`
>
> `:=` enables dynamic naming of variables with tidy evaluation (used
> when there is only variable to name in the function)

```{r name_assignment_soln}
# Use the Walrus operator and syntax similar to the glue package
group_mean_name <- function(data, grouping_variable, summary_column){
  data %>%
    group_by({{ grouping_variable }}) %>%
    summarise("mean_{{ summary_column }}" := mean({{ summary_column }}, na.rm = T))
}

group_mean_name(gapminder, country, lifeExp)
```

### **3. Indirection with string inputs**

Sometimes you will need to interpret strings as data variable names.
This may be necessary if the name of the required data variable is read
in from a separate file or from user input, and held within an
environment variable.

Enabling your function to accept string inputs also avoids perpetuating
data masking into your function (which may be what you want if the
function will be used by programmers not familiar with data masking!).

> #### SOLUTION - use data sub-setting explicitly with: `.data[[ var ]]`
>
> When you have a character env-variable and you wish to input this into
> a function, you need to index into the `.data` pronoun with `[[`.

-   Note that `.data` is not a data frame; it's a special construct, a
    pronoun, within the tidyverse that allows you to access the current
    variables either directly, with `.data$x` or indirectly with
    `.data[[var]]` (in a grouped dataset it represents the subset for
    the current group). Don't expect other functions to work with it.

> If you want to assign a name using string inputs (either function
> arguments or environment variables) then use `:=` + `{ }` (***single
> curly brackets***)

```{r indirection_with_strings}
group_mean <- function(data, grouping_variable, summary_column){
  data %>%
    group_by(.data[[grouping_variable]]) %>%
    summarise("mean_{summary_column}" := mean(.data[[summary_column]], na.rm = T))
}

#It works!
group_mean(gapminder, "year", "lifeExp")
```

------------------------------------------------------------------------

## Tidy Selection: Allow arbitrary selection of variables

You might want to create a function that accepts sets of (not
necessarily sequential) columns as input (similar to how tidyverse
`select()` functions work).

In simple cases the solution is obtained with curly-curly `{{ }}`:

> #### Solution 1:
>
> [Unquoted arguments + tidy-select]{.underline}: Embrace the function
> input variables using `{{ }}` when using tidyverse functions with
> tidy-select:

```{r slice_example}
slice_fun_tidy <- function(data, select_columns, rows) {
  data %>%
    ## select() uses <tidy-select>
    select({{ select_columns }}) %>%
    slice(rows)
}

slice_fun_tidy(gapminder, lifeExp:gdpPercap, 1)
slice_fun_tidy(gapminder, where(is.double), 1)

```

> #### Solution 2:
>
> [Unquoted arguments more generally]{.underline}: Use across() and
> `{{ }}` to combine the function input variables when using tidyverse
> functions WITHOUT tidy-select:

```{r slice_example}
slice_fun_untidy <- function(data, grouping_variable, rows) {
  data %>%
    ## group_by() doesn't use <tidy-select>
    group_by(across({{ grouping_variable }})) %>%
    slice(rows)
}

slice_fun_untidy(gapminder, c(country,year), 1)

```

> #### Solution 3:
>
> [Quoted arguments]{.underline}: Use `all_of()/any_of()` when the input
> is a list of strings giving the variable names (or you don't want to
> propagate data masking).
>
> -   `all_of()` acts on all of the columns listed (they must all
>     exist).
> -   `any_of()` acts on all of the columns listed, if they exist, and
>     ignores missing ones

```{r}
slice_stringinputs <- function(data, columns, rows) {
  data %>%
    select(all_of(columns)) %>%
    slice(rows)
}

slice_stringinputs(gapminder, c("country", "year"), 1)

## Again you need 'across' when using non-<tidy-select> functions
slice_stringinputs <- function(data, columns, rows) {
  data %>%
    group_by(across(all_of(columns))) %>%
    slice(rows)
}
slice_stringinputs(gapminder, c("country", "year"), 1)

```

## What about ggplot? - it works the same way! See `Extras.Rmd`

## Do you even need Tidy Eval? - see `Extras.Rmd`

## References:

Type `?dplyr_data_masking` in your R console for the R documentation on
data masking and examples.

Lost of **super documentation** in the rlang website - especially these
two pages:

-   <https://rlang.r-lib.org/reference/topic-data-mask.html>

-   <https://rlang.r-lib.org/reference/topic-data-mask-programming.html>

YouTube videos by Lionel Henry (who wrote rlang):

-   <https://www.youtube.com/watch?v=Tmy8s01t-Ao>

-   <https://www.youtube.com/watch?v=SW_TUkPXSmA>

And this git repo by Dr Tiffany Timbers (this tutorial and worksheet is
modeled on hers!) **#Attribution!**

-   <https://github.com/ttimbers/prog-tidyverse-functions>

Overview of Standard and Non-standard Evaluation (R focused)

-   <https://www.brodieg.com/2020/05/05/on-nse/>

------------------------------------------------------------------------
