---
title: "Tidyverse programming tutorial"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
```

## Tidyverse is easy to work with!

Let's suppose we want to calculate the mean life expectancy per
continent per year for the Gapminder dataset:

```{r contrived_example}
meanLE_year <- gapminder %>% 
  group_by(year) %>% 
  summarise(meanLE = mean(lifeExp, na.rm = T))

meanLE_continent <- gapminder %>% 
  group_by(continent) %>% 
  summarise(meanLE = mean(lifeExp, na.rm = T))

meanLE_country <- gapminder %>% 
  group_by(country) %>% 
  summarise(meanLE = mean(lifeExp, na.rm = T))


ggplot(meanLE_year, aes(x = year, y = meanLE)) + geom_point() + geom_smooth(method = "lm")

ggplot(meanLE_continent, aes(x = continent, y = meanLE)) + geom_point() 

ggplot(meanLE_country, aes(x = country, y = meanLE)) + geom_point() 

```

Simple - right!? Once we specify the dataset we are working with we can
directly type the names of the columns and variables in the dataset we
wish without have to enclose them in quotation marks or having to define
them relative to the dataset itself.

## But - there's a catch.. the problem:

What happens if we want to turn the above code into a function?  - which
we might want to do if we wanted to get and plot the means of population
size and gdpPercap as well as lifeExp across continents and years. Or if
want to get the means by countries/years instead.

```{r try_function, error=TRUE}
# Place code within function

gap_means <- function(data, grouping_variable, summary_column){
  data %>% 
  group_by(grouping_variable) %>% 
  summarise(meanLE = mean(summary_column, na.rm = T))
}

#Does it work?
gap_means(gapminder, year, lifeExp)

#What about if we place the arguments in quotes?
gap_means(gapminder, "year", "lifeExp")

#Another example - What if tried to build a generic plotting function for all of the above dataset?
plot_fun <- function(data, x, y){
  data %>% 
  ggplot(aes(x = x, y = y)) +
  geom_point() 
}

plot(meanLE_continent, x = continent, y = meanLE)
plot(meanLE_continent, x = "continent", y = "meanLE")

```

## NSE = Non-standard evaluation ≈ Tidy evaluation

What is NSE? Very roughly, it is to programmatically modify an
expression or its meaning after it is issued but before it is executed.
You can think of an "expression" an R command you might type at the
prompt or in an R script.

*Not many programming languages give the programmer the power to
implement, use, and abuse Non-Standard Evaluation  - but R does!*

And not just in Tidyverse - ALSO in base R.

```{r NSE_baseR}
subset(mtcars, hp > 250) # this works 
                         # subset delays the evalution of hp > 250 before it is run 
                         # changes its meaning/environment so it relates to mtcars 
mtcars[hp > 250, ] # Does this work?
                   # NO! hp is not defined in our environment

#Other base R NSE functions include:
transform()
lm()
with()
#and the old school (for those of us who learnt R way way before the tidyverse!):
attach()
```

Much more to learn about NSE!! See list of references at end.

#### Tidy evaluation has two forms: Data Masking and Tidy Selection

Tidy evaluation is a special type of non-standard evaluation used
throughout the {tidyverse}.

-   `arrange()`, `count()`, `filter()`, `group_by()`, `mutate()`, and
    `summarise()` use **data masking** so that you can use data
    variables as if they were variables in the environment (i.e. you
    write `my_variable` not `df$myvariable`).

-   `across()`, `relocate()`, `rename()`, `select()`, and `pull()` use
    **tidy selection** so you can easily choose variables based on their
    position, name, or type (e.g. `starts_with("x")` or `is.numeric`).

To determine whether a function argument uses data masking or tidy
selection, look at the documentation: in the arguments list, you'll see
`<data-masking>` or `<tidy-select>`. Let's look at a few now!

```{r}
?dplyr::arrange
```

```{r}
?dplyr::rename
```

Data masking and tidy selection make interactive data exploration fast
and fluid, but they add some new challenges when you attempt to use them
indirectly such as in a for loop or a function. This talk will show you
how to overcome those challenges.

Before we get started, let's load the tidyverse packages:

```{r load tidyverse, message = FALSE}
library(tidyverse)
```

It turns out this won't work! Why not? The unquoted column names trip R
up here! Let's learn more about why, and then what we can do to handle
this!

## Data masking

Data masking makes data manipulation faster because it requires less
typing. In most (but not all) base R functions you need to refer to
variables with `$`, leading to code that repeats the name of the data
frame many times:

```{r, results = FALSE}
starwars[starwars$homeworld == "Naboo" & starwars$species == "Human", ]
```

The dplyr equivalent of this code is more concise because data masking
allows you to need to type `starwars` once:

```{r, results = FALSE}
starwars %>% filter(homeworld == "Naboo", species == "Human")
```

### Data- and env-variables

The key idea behind data masking is that it blurs the line between the
two different meanings of the word "variable":

-   **env-variables** are "programming" variables that live in an
    environment. They are usually created with `<-`.

-   **data-variables** are "statistical" variables that live in a data
    frame. They usually come from data files (e.g. `.csv`, `.xls`), or
    are created manipulating existing variables.

To make those definitions a little more concrete, take this piece of
code:

```{r}
df <- data.frame(x = runif(3), y = runif(3))
df$x
```

It creates a env-variable, `df`, that contains two data-variables, `x`
and `y`. Then it extracts the data-variable `x` out of the env-variable
`df` using `$`.

I think this blurring of the meaning of "variable" is a really nice
feature for interactive data analysis because it allows you to refer to
data-vars as is, without any prefix. And this seems to be fairly
intuitive since many newer R users will attempt to write
`diamonds[x == 0 | y == 0, ]`.

Unfortunately, this benefit does not come for free. When you start to
program with these tools, you're going to have to grapple with the
distinction!

### Indirection

The main challenge of programming with functions that use data masking
arises when you introduce some indirection, i.e. when you want to get
the data-variable from an env-variable instead of directly typing the
data-variable's name. There are two main cases:

-   When you have the data-variable in a function argument (i.e. an
    env-variable that holds a promise[^1]), you need to **embrace** the
    argument by surrounding it in doubled braces, like
    `filter(df, {{ var }})`.

    The following function uses embracing to create a wrapper around
    `summarise()` that computes the minimum and maximum values of a
    variable, as well as the number of observations that were
    summarised:

[^1]: In R, arguments are lazily evaluated which means that until you
    attempt to use, they don't hold a value, just a **promise** that
    describes how to compute the value. You can learn more at
    <https://adv-r.hadley.nz/functions.html#lazy-evaluation>

```{r}
group_means <- function(data, grouping_column, column_to_summarize) {
  data %>% 
    group_by({{ grouping_column }}) %>% 
    summarise(column_mean = mean({{ column_to_summarize }}))
}
```

```{r}
group_means(starwars, species, height)
```

-   When you have an env-variable that is a character vector, you need
    to index into the `.data` pronoun with `[[`, like
    `summarise(df, mean = mean(.data[[var]]))`.

    The following example uses `.data` to count the number of unique
    values in each variable of `mtcars`:

    ```{r}
    group_means <- function(data, grouping_column, column_to_summarize) {
      data %>% 
        group_by(.data[[grouping_column]]) %>% 
        summarise(column_mean = mean(.data[[column_to_summarize]]))
    }
    ```

    ```{r}
    group_means(starwars, "species", "height")
    ```

    Note that `.data` is not a data frame; it's a special construct, a
    pronoun, that allows you to access the current variables either
    directly, with `.data$x` or indirectly with `.data[[var]]`. Don't
    expect other functions to work with it.

## Tidy selection

Data masking makes it easy to compute on values within a dataset. Tidy
selection is a complementary tool that makes it easy to work with the
columns of a dataset.

### The tidyselect DSL

Underneath all functions that use tidy selection is the
[tidyselect](https://tidyselect.r-lib.org/) package. It provides a
miniature domain specific language that makes it easy to select columns
by name, position, or type. For example:

-   `select(df, 1)` selects the first column; `select(df, last_col())`
    selects the last column.

-   `select(df, c(a, b, c))` selects columns `a`, `b`, and `c`.

-   `select(df, starts_with("a"))` selects all columns whose name starts
    with "a"; `select(df, ends_with("z"))` selects all columns whose
    name ends with "z".

-   `select(df, where(is.numeric))` selects all numeric columns.

You can see more details in `?dplyr_tidy_select`.

### Indirection

As with data masking, tidy selection makes a common task easier at the
cost of making a less common task harder. When you want to use tidy
select indirectly with the column specification stored in an
intermediate variable, you'll need to learn some new tools. Again, there
are two forms of indirection:

-   When you have the data-variable in an env-variable that is a
    function argument, you use the same technique as data masking: you
    **embrace** the argument by surrounding it in doubled braces.

    The following function subsets a data frame using numerical row
    indexing and numbers or names for column indexing:

    ```{r, results = FALSE}
    rectangle <- function(data, columns, row_start, row_end) {
      data %>% 
        select({{ columns }}) %>% 
        slice(row_start:row_end)
    }
    ```

    ```{r}
    rectangle(starwars, height:eye_color, 1, 5)
    ```

-   When you have an env-variable that is a character vector, you need
    to use `all_of()` or `any_of()` depending on whether you want the
    function to error if a variable is not found.

    The following code uses `all_of()` to select all of the variables
    found in a character vector; then `!` plus `all_of()` to select all
    of the variables *not* found in a character vector:

    ```{r}
    rectangle <- function(data, columns, row_start, row_end) {
          data %>% 
            select(all_of(columns)) %>% 
            slice(row_start:row_end)
        }
    ```

    ```{r}
    vars <- c("height", "mass")
    rectangle(starwars, vars, 1, 5)
    ```

## The Walrus operator

`:=` is named the "Walrus operator" and is needed in some cases when
assigning values with tidy evaluation. For example, what if we wanted to
take our `group_means` function from earlier and improved it so that the
column name in the summarized data frame was the same column name as in
the original data frame?

We might then try something like this:

```{r, error=TRUE}
group_means <- function(data, grouping_column, column_to_summarize) {
  data %>% 
    group_by({{ grouping_column }}) %>% 
    summarise({{ column_to_summarize }} = mean({{ column_to_summarize }}))
}
```

But it looks like we cannot even define that function! Anytime we want
to use indirection with a column name during assignment, we need to use
the `:=` to make tidy evaluation work correctly:

```{r}
group_means <- function(data, grouping_column, column_to_summarize) {
  data %>% 
    group_by({{ grouping_column }}) %>% 
    summarise({{ column_to_summarize }} := mean({{ column_to_summarize }}))
}
```

Now we can define, and use our function!

```{r}
group_means(starwars, species, height)
```

We can even go further and combine our indirected column name with a
string to improve the column name further:

```{r}
group_means <- function(data, grouping_column, column_to_summarize) {
  data %>% 
    group_by({{ grouping_column }}) %>% 
    summarise("mean_{{ column_to_summarize }}" := mean({{ column_to_summarize }}))
}
```

```{r}
group_means(starwars, species, height)
```

## Using tidy evaluation in other {tidyverse} package functions

Finally, we can apply this same logic to other {tidyverse} package
functions (so far we have just focused on {dplyr}), such as {ggplot2}
functions!

```{r}
scatter_plot <- function(data_frame, x_axis, y_axis) {
  ggplot(data_frame, aes(y = {{ y_axis }}, x = {{ x_axis }})) +
    geom_point(alpha = 0.5)
}
```

```{r}
scatter_plot(starwars, height, mass)
```

> If you'd like to learn more about the underlying theory, or precisely
> how it's different from non-standard evaluation, we recommend that you
> read the Metaprogramming chapters in [*Advanced
> R*](https://adv-r.hadley.nz).
