---
title: "Tidyverse programming tutorial"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
```

## ♥ Tidyverse = super easy to work with! ♥

Let's suppose we want to calculate the mean life expectancy per
continent or per year for the
[Gapminder](https://www.gapminder.org/data/documentation/ "Description of data from gapminder.org"){.uri}
dataset:

```{r contrived_example}
# Check what the dataset looks like
head(gapminder)

# Use tidyverse to calculate grouped means
gapminder %>%
  group_by(year) %>%
  summarise(meanLE = mean(lifeExp, na.rm = T))
gapminder %>% 
  group_by(continent) %>% 
  summarise(meanLE = mean(lifeExp, na.rm = T))
```

Simple! Once we specify the dataset we are working with we can directly
type the names of the columns and variables in the dataset, without
having to enclose them in quotation marks or having to define them
relative to the dataset itself.

## But there's a problem! Trouble in paradise?

In the example above we calculated the same metric for two different
grouping variables by repeating the code and changing one element. In
real examples of course, the code is likely to be much longer. This
isn't good coding practice. When we find ourselves repeating code we
should look at using some kind of iteration method or writing a bespoke
function.

Also, we might want to get the mean of population size and GDP per
capita as well as life expectancy across continents and years. Or we
might want to calculate it by countries AND years. We shouldn't repeat
the above code that many times! We need to write re-usable code.

SO let's see what happens when we try to turn the above code into a
function:

```{r try_function}
# Place code within function

group_mean <- function(data, grouping_variable, summary_column){
  data %>% 
  group_by(grouping_variable) %>% 
  summarise(mean = mean(summary_column, na.rm = T))
}

# Does it work?
group_mean(gapminder, year, lifeExp)

# Oh no! R does not understand what we mean. 
# What about if we place the arguments in quotes?
group_mean(gapminder, "year", "lifeExp")
# Still no. What now?

```






```

These approaches all fail because the function syntax confuses
environment variables with data variables - which happens because
R/Tidyverse uses Non-standard Evaluation

------------------------------------------------------------------------

## NSE = Non-standard evaluation (VERY VERY briefly!)

What is NSE? Very roughly, it is to "*programmatically modify an
expression or its meaning after it is issued but before it is
executed*". (You can think of an expression as an R command you might
type at the prompt or in an R script).

*Not many programming languages give the programmer the power to
implement, use, and abuse Non-Standard Evaluation - but R does!*

And not just in Tidyverse - ALSO in base R!

```{r NSE_baseR}
subset(gapminder, continent=="Asia") # this works because subset delays the 
                                     # evaluation of continent=="Asia" until it 
                                     # is run when it changes its 
                                     # meaning/environment so it relates to gapminder
gapminder[continent=="Asia", ] # Does this work?
                               # NO! continent is not defined in our environment

#Other base R NSE functions include:
# with(); transform(); lm()
# and the old school (for those of us who learnt R way before the tidyverse!):
# attach()

```

Much more to learn about NSE!! See list of references at end.

#### ≈ Tidy Evaluation (NSE in the Tidyverse; implemented in the rlang package) has two forms:

**Data Masking** - allows you to use data variables as if they were
variables in the environment/ objects in the workspace --\> i.e. lets
you write `my_variable` not `dataset$myvariable`)

-   `arrange()`, `count()`, `filter()`, `group_by()`, `mutate()`, and
    `summarise()`

Data masking makes data manipulation faster because it requires less
typing (cf with base R):

```{r data_masking, results = FALSE}
#Comapre Base R
gapminder[gapminder$continent == "Europe" & gapminder$pop < 1000000, ]

#With Tidyverse
gapminder %>% filter(continent == "Europe", pop < 1000000)
```

**Tidy Selection** - allows you to easily choose variables based on
their position, name, or type (e.g. `starts_with("x")` or `is.numeric`).

-   `across()`, `relocate()`, `rename()`, `select()`, and `pull()`

> Tidyverse functions are built for interactivity! They make interactive
> data exploration fast and fluid (and fun!)
>
> But they add some new challenges when you attempt to use them
> indirectly such as in a for loop or a function.

------------------------------------------------------------------------

## Data Masking Problems and Solutions

### **1. Unexpected masking**

The key idea behind data masking is that it blurs the line between the
two different meanings of the word *variable*:

-   **environment variables** are programming variables that live in an
    environment. They are usually created with `<-`.

-   **data variables** are statistical variables that live in a data
    frame (come from data files or are created by manipulating variables
    within the data frame).

```{r unexp_masking}
# Example of Unexpected masking
# An object in your workspace (which you want to use) has the same name as a dataset column
n <- 100
data1 <- data.frame(x = 1)
data2 <- data.frame(x = 1, n = 2)

data1 %>% mutate(y = x / n) %>% pull(y)
data2 %>% mutate(y = x / n) %>% pull(y)

```

> #### SOLUTION - Be explicit in your code
>
> Use the `.data$` or `.env$` pronouns to explicitly associate variables
> to data frames or the global environment:

```{r unexp_mask_soln}

#Specify .data and .end pronouns
data1 %>% mutate(y = .data$x / .env$n) %>% pull(y)
data2 %>% mutate(y = .data$x / .env$n) %>% pull(y)
```

### **2. Indirection**

Occurs when data-variables can't get through functions/loops because the
function expects an environment variable (but you provide a data
variable in the call):

```{r indrection}
# Example of indirection = function we tried to run above - DOESN'T WORK
gap_mean <- function(data, grouping_variable, summary_column){
  data %>% 
  group_by(grouping_variable) %>% 
  summarise(mean = mean(summary_column, na.rm = T))
}

#It doesn't work
gap_mean(gapminder, year, lifeExp)
```

Inside the function, `grouping_variable` and `summary_column` are
environment variables. But when you call the function `year` and
`lifeExp` are data variables.

Data-masked code needs to be *delayed* and *transported* to the data
context. Behind the scenes, dplyr verbs achieve this by capturing the
blueprint of your code, and resuming its evaluation inside the data
mask. The example above fails because `group_by()` is capturing the
wrong piece of blueprint.

> #### SOLUTION - use the curly-curly operator `{{ }}`

> The curly-curly `{{ }}` operator allows you to *tunnel* a data
> variable through a function. It acts as a wrapper for lower level
> operators (`enquo()` and `!!`) that function to **quote-and-unquote**
> the expression pattern:
>
> `enquo()` - delays the interpretation of code and captures its
> blueprint and `!!` - which modifies those blueprints.

```{r indrection_soln}
gap_mean <- function(data, grouping_variable, summary_column){
  data %>% 
  group_by({{ grouping_variable }}) %>% 
  summarise(mean({{ summary_column }}, na.rm = T))
}

#It works!
gap_mean(gapminder, year, lifeExp)
gap_mean(gapminder, continent, pop)

```

### 3. Naming of variables in functions

> #### Solution - use the walrus operator `:=`
>
> `:=` enables dynamic naming of variables with tidy evaluation (used
> when there is only variable to name in the function)

```{r, name_assignment}
# Use the Walrus operator and syntax similar to the glue package
gap_mean_name <- function(data, grouping_variable, summary_column){
  data %>% 
  group_by({{ grouping_variable }}) %>% 
  summarise("mean_{{ summary_column }}" := mean({{ summary_column }}, na.rm = T))
}

gap_mean_name(gapminder, country, lifeExp)
```

### 4. Indirection with string inputs?

> #### SOLUTION A - use data sub-setting explicitly with: `.data[[ var ]]`
>
> When you have a character env-variable and you wish to input this into
> a function, you need to index into the `.data` pronoun with `[[`.
>
> If you want to assign name then use `:=` + `{ }` (***single curly
> brackets***)

```{r indirection_with_strings}
gap_means <- function(data, grouping_variable, summary_column){
  data %>% 
  group_by(.data[[grouping_variable]]) %>% 
  summarise("mean_{summary_column}" := mean(.data[[summary_column]], na.rm = T))
}

#It works!
gap_means(gapminder, "year", "lifeExp")
```

-   Note that `.data` is not a data frame; it's a special construct, a
    pronoun, within the tidyverse that allows you to access the current
    variables either directly, with `.data$x` or indirectly with
    `.data[[var]]` (it represents the subset for the current group).
    Don't expect other functions to work with it.

-   Note also that using this syntax avoids creating new bespoke
    functions that ARE ALSO DATA MASKING functions! This can be good
    when your code will be used by non-R users or in production.

------------------------------------------------------------------------

## Tidy Selection: Problems and Solutions

As with data masking, tidy selection makes a common task easier at the
cost of making a less common task harder. The solutions are similar to
those described:

1.  Embrace the function input vars using `{{ }}` - when the input is
    not a string input

    ```{r slice_example}
    slice_fun <- function(data, grouping_variable, columns, rows) {
      data %>% 
        group_by({{ grouping_variable }}) %>% 
        select({{ columns }}) %>% 
        slice(rows)
    }
    slice_fun(gapminder, continent, lifeExp:gdpPercap, 1)
    ```

-   Alternatively - if you have string input (or don't want to propagate
    data masking) then use `.data[[ ]]`:

    ```{r}
    slice_character_subset <- function(data, column, row_start, row_end) {
      data %>% 
        select(.data[[column]]) %>% 
        slice(row_start:row_end)
    }
    slice_character_subset(gapminder, "pop", 1, 10)
    ```

## Using tidy evaluation in other {tidyverse} package functions: ggplot

Finally, we can apply this same logic to other {tidyverse} package
functions.

-   either embrace `{{ }}` the data variable names

-   or subset the data using `.data [[ ]]` and input strings to the
    function call

```{r}
scatter_plot <- function(data_frame, x_axis, y_axis ) {
  ggplot(data_frame, aes(x = {{ x_axis }}, y = {{ y_axis }})) +
    geom_point(alpha = 0.5)
}

gapminder %>% 
  gap_mean_name(year, lifeExp) %>% 
  scatter_plot(year, mean_lifeExp)

gapminder %>% 
  gap_mean_name(continent, gdpPercap) %>% 
  scatter_plot(continent, mean_gdpPercap)


scatter_plot <- function(data_frame, x_axis, y_axis ) {
  ggplot(data_frame, aes(x = .data[[x_axis]], y = .data[[y_axis]])) +
    geom_point(alpha = 0.5)
}

gapminder %>% 
  gap_mean_name(continent, gdpPercap) %>% 
  scatter_plot("continent", "mean_gdpPercap")
```

## What about with multiple data variables? - see `Extras.Rmd`

## Do you even need Tidy Eval? - see `Extras.Rmd`

## References:

Type `?dplyr_data_masking` in your R console for the R documentation on
data masking and examples.

Lost of **super documentation** in the rlang website - especially these
two pages:

-   <https://rlang.r-lib.org/reference/topic-data-mask.html>

-   <https://rlang.r-lib.org/reference/topic-data-mask-programming.html>

YouTube videos by Lionel Henry (who wrote rlang):

-   <https://www.youtube.com/watch?v=Tmy8s01t-Ao>

-   <https://www.youtube.com/watch?v=SW_TUkPXSmA>

And this git repo by Dr Tiffany Timbers (this tutorial and worksheet is
modeled on hers!) #Attribution

-   <https://github.com/ttimbers/prog-tidyverse-functions>

Overview of Standard and Non-standard Evaluation (R focused)

-   <https://www.brodieg.com/2020/05/05/on-nse/>

------------------------------------------------------------------------
